Edge sign prediction in general graphs and Correlation Clustering

### 2017-10-02 troll/algs_batch.tex
explain better the first troll batch algo

Given $G(Y) =(V,E(Y))$, in the batch setting we have at our disposal a training set \trainset{} of
labeled edges from $E(Y)$. We want to build a predictive model for the labels of the remaining
edges. We present two algorithms to do so, which both compute estimates of the all parameters $p$
and $q$ of our generative model. They differ in the approximation guarantees they provide, and in
the class of graphs to which they apply. The first algorithm runs on the graph $G$ and estimates
locally the parameters by their empirical means in the training set, which under some density
assumptions are showed to concentrate around their true values. The second algorithm, on the other
hand, exploits the reduced graph $G''$ and computes a maximum likelihood estimation of the
parameters through a global label propagation approach, without making any density assumption.

@@

Our first algorithm is an approximation to the Bayes optimal predictor $y^*(u,v)$. Let us denote by
$\htr(u)$ and $\hun(u)$ the trollness and the untrustworthiness of node $u$ when both are computed
on the subgraph induced by the training edges. We now design and analyze an edge classifier of the
form

\begin{equation}
  \label{eq:predictor}
  \sgn\Big(\big(1-\htr(u)\big) + \big(1-\hun(v)\big) -\tau - \tfrac{1}{2}\Big)~,
\end{equation}

where $\tau \ge 0$ is the only parameter to be trained. Despite its simplicity, this classifier
works reasonably well in practice, as demonstrated by our experiments (see \autoref{s:exp}).
Moreover, unlike previous edge sign prediction methods for directed graphs, our classifier comes
with a rigorous theoretical motivation, since it approximates the Bayes optimal classifier
$y^*(u,v)$ with respect to the generative model defined in \autoref{s:gen}. It is important to point
out that when we use $1-\htr(u)$ and $1-\hun(v)$ to estimate $p_u$ and $q_v$, an additive bias shows
up due to \eqref{e:pout} and \eqref{e:pin}. This motivates the need of a threshold parameter $\tau$
to cancel this bias. Yet, the presence of a prior distribution $\mu(p,q)$ ensures that this bias is
the same for all edges $(u,v) \in E$.

\iffalse

@@

\textbf{4.} Any remaining non-sampled edge $(u,v)$ is predicted as $ \yhat(u,v) =
\sgn\big(\hdeltaout(u) + \hdeltain(v) - \tfrac{1}{2} - \tauhat \big) $.
\fi

Our algorithm works under the assumption that for given parameters $Q$ (a positive integer) and
$\alpha \in (0,1)$, there exists a set\footnote{$E_L$ is needed to find an estimate $\tauhat$ of
$\tau$ in~(\ref{eq:predictor}), as done in Step~3 of the algorithm. Any undirected matching of $G$ of
size $O(\log|V|)$ can be used, obtained for instance by the blossom algorithm~\autocite{matching65}.
In practice, however, we never computed $E_L$, and estimated $\tau$
on the entire training set \trainset{}.} $E_L \ss E$ of size $\tfrac{2Q}{\alpha}$ where each vertex
$u \in V$ appearing as an endpoint of some edge in $E_L$ occurs at most once as origin ---\ie{}
$(u,v)$--- and at most once as destination ---\ie{} $(v,u)$. Moreover, we assume \trainset{} has
been drawn from $E$ at random {\em without} replacement, with $m = |E_0| = \alpha\,|E|$. The
algorithm performs the following steps:

\begin{enumerate}[label=\textbf{\arabic*.}]
	\item For each $v \in V$, let $\hun(v) = \hdin^-(v)/\hdin(v)$, \ie{} the fraction of
	  negative edges found in $\Nin(v) \cap \trainset{}$.

	\item For each $u \in V$, let $\htr(u) = \hdout^-(u)/\hdout(u)$, \ie{} the fraction of
	  negative edges found in $\Nout(u) \cap \trainset{}$.

	\item Let $\tauhat$ be the fraction of positive edges in $E_L\cap \trainset{}$.

	\item Any remaining edge $(u,v) \in E\setminus \trainset{}$ is predicted as
		$$
		\yhat(u,v) = \sgn\Big(\big(1-\htr(u)\big) + \big(1-\hun(v)\big) -\tauhat - \tfrac{1}{2}\Big)
		$$
\end{enumerate}

The next result shows that the above algorithm can approximate the Bayes optimal predictor on nodes
whose in-degree and out-degree is not too small.

\begin{theorem}
  \label{t:active}
  Let $G(Y) = (V,E(Y))$ be a directed graph with labels on the edges generated according to the
  model in \autoref{s:gen}.
  If the algorithm is run with parameter $Q = \Omega(\log|V|)$, and $\alpha \in (0,1)$ such that the
  above assumptions are satisfied, then $\yhat(u,v) = y^*(u,v)$ holds with high probability
  simultaneously for all test edges $(u,v) \in E$ such that $\dout(u),\din(v) = \Omega(\log|V|)$, and
  $\eta(u,v) = \Pr(\yuv=1)$ is bounded away from $\tfrac{1}{2}$.
\end{theorem}

While we defer the full proof of \autoref{t:active} to the additional material at the end of this
chapter (\autoref{ssec:troll_proof_bayes}), here we give a sketch of the method used, focusing on
parameter $p$. First, by \autoref{l:bins} on negatively associated random variables (stated
\vpageref{l:bins}), we show that with our choice of $Q$, for $\theta=\frac{2Q}{\alpha}$ and for any
$u$ having enough out neighbors (\ie{} $\NNout(u) \ge \theta$), at least $Q$ edges of $\Nout(u)$
are in the training set with high probability.
We use these $Q$ edges to compute $1-\htr(u)$, the empirical probability of drawing a $+1$-labeled
edge from $\Nout(i)$. We show that it is close to the real probability, which according to equation
\eqref{e:pout} is equal to $\frac{1}{2}\,\left(p_u + \qbar_u\right)$. $\qbar_u$ being a sample mean
of i.i.d.~$[0,1]$-valued random variables independently drawn from the prior marginal $\int_0^1
\mu\big(p,\cdot\big) dp$, it concentrates around its expectation $\mu_q$.

By following the same reasoning on the parameters $q$ and using an union bound, we then show that
$\left(1-\htr(u)\right) + \left(1-\hun(v)\right) - \frac{\mu_p+\mu_q}{2}$ is close to
$\frac{p_u+q_v}{2}$ simultaneously for all edges. Another set of concentration arguments then show
that our estimate $\tauhat$ is close to $\frac{\mu_p+\mu_q}{2}$. Combining these two facts ends the
proof.

@@

The approach leading to \autoref{t:active} needs the graph to be sufficiently dense. At first sight,
having $Q = \Omega(\log|V|)$ training edges per nodes does not sound that like such a stringent
requirement. Consider for instance Facebook ---which is neither signed nor directed though. It has
two billion
users~\bottomfootnote{\url{https://investor.fb.com/investor-news/press-release-details/2017/Facebook-Reports-Second-Quarter-2017-Results/}},
each of them having $155\approx 7.2\log|V|$ friends on average~\autocite{facebookFriend16}. However,
the constant hidden in the $\Omega$ notation, and needed to guarantee a good behavior, is much
larger than $7.2$. We will nonetheless see in the experiments that we can still apply the resulting
algorithm (we call it \usrule{}) with satisfying results. Additionally, and in order to address this
density limitation, we now introduce a second method based on label propagation.

@@

We now proceed to solve \eqref{e:quadratic} via label propagation~\autocite{LabelPropa03} on the
graph $G''$ obtained through the $G \rightarrow G''$ reduction of \autoref{s:prel}.
However, because of the presence of negative edge weights in $G''$, we first have to symmetrize\footnote{%
While we note here that such linear transformation of the variables does not change the problem, we
provide more details in \autoref{ssec:troll_lprop_weights} of the supplementary material.} variables
$p_i, q_i$ and $\yuv$ so as they all lie in the interval $[-1,+1]$.
After this step, one can see that, once we get back to the original variables, label propagation
computes the harmonic solution minimizing the function

>>>>

Given $G(Y) =(V,E(Y))$, in the batch setting we have at our disposal a training set \trainset{} of
labeled edges from $E(Y)$. Formally, we have a training set $\trainset
= \big((u_1,v_1),y_{u_1,v_1}), ((u_2,v_2),y_{u_2,v_2}), \ldots, ((u_m,v_m),y_{u_m,v_m}\big)$, that
has been drawn from $E\times Y$ \uar{} with replacement, with
$m = |E_0|$. Then a reasonable approach to approximate $y^*(u,v)$ would be to resort to a maximum
 We want to build a predictive model for the labels of the remaining
edges. We present two algorithms to do so, which both compute estimates of the all parameters $p$
and $q$ of our generative model. They differ in the approximation guarantees they provide, and in
the class of graphs to which they apply. The first algorithm runs on the graph $G$ and estimates
locally the parameters by their empirical means in the training set, which under some density
assumptions are showed to concentrate around their true values. The second algorithm, on the other
hand, exploits the reduced graph $G''$ and computes a maximum likelihood estimation of the
parameters through a global label propagation approach, without making any density assumption.

@@

Our first algorithm is an approximation to the Bayes optimal predictor $y^*(u,v)$. Let us denote by
$\htr(u)$ and $\hun(u)$ the trollness and the untrustworthiness of node $u$ when both are computed
on the subgraph induced by the training edges.
Recall that the Bayes optimal predictor classifies an edge \euv{} using the following rule:
$y^*(u,v) = \sgn\left( \frac{p_u+q_v}{2} - \frac{1}{2} \right)$. Our approximation thus consists in
using the quantities $\htr(u)$ and $\hun(u)$ to estimate $\frac{p_u+q_v}{2}$. This results in the
following rule:
\begin{equation}
  \label{eq:predictor}
  \sgn\Big(\big(1-\htr(u)\big) + \big(1-\hun(v)\big) - \tau - \tfrac{1}{2}\Big)~,
\end{equation}
where $\tau \ge 0$ is the only parameter to be trained. We now give an intuition to justify this
equation, and defer the technical arguments to end of the chapter,
\vpageref{ssec:troll_proof_bayes}, where we will formalize what it means for a quantity to be
\enquote{close} to another. Note first that $1-\htr(u) = \frac{\hdout^+(u)}{\hdout(u)}$ is the
empirical mean\footnote{The hat symbol denote quantities computed solely on the training set.} of
the probability of a random edge outgoing from $u$ to be positive, and is therefore \enquote{close}
to $\frac{1}{2}\,\left(p_u + \qbar_u\right)$ according to \eqref{e:pout}. By the same reasoning,
$1-\hun(v) \approx \frac{1}{2}\,\left(q_v + \pbar_v\right)$ in accordance with \eqref{e:pin}. At
this stage, $\left( 1 - \htr(u) \right) + \left( 1 - \hun(v) \right) - \frac{p_u+q_v}{2} \approx
\lhalf \left( \pbar_v + \qbar_u \right)$. Since $\pbar_v$ is a sample mean of i.i.d.~$[0,1]$-valued
random variables independently drawn from the prior marginal $\int_0^1 \mu\big(\cdot, q\big) dq$, it
concentrates around its expectation $\mu_p$. Likewise, $\qbar_u \approx \mu_q$. Now we seek an
estimation of $ \lhalf \left( \pbar_v + \qbar_u \right) \approx \lhalf \left( \mu_p + \mu_q
\right)$. For that, we compute the expected number of positive edges in the graph, which, as in the
previous section, is the expected value of the random variable $Z = \sum_{\euv \in E} \Ind{\yuv =
+1}$. Let $V_{\mathrm{out}}$ be the subset of node of $V$ with at least one outgoing edge, and
define similarly $V_{\mathrm{in}}$ as $\theset{v\in V}{\din(v) > 0}$. According to our generative
model, $\E\left(Z \right) = \sum_{\euv \in E} \frac{p_u+q_v}{2} = \sum_{u \in V_{\mathrm{out}}}
\left( \sum_{v \in \NNout(u)} \frac{p_u+q_v}{2} \right)$. Observe that
\begin{align*}
  \sum_{u \in V_{\mathrm{out}}} \left( \sum_{v \in \NNout(u)} p_u \right) &=
  \sum_{v \in V_{\mathrm{in}}} \left( \sum_{u \in \NNin(v)} p_u \right) =
  \sum_{v \in V_{\mathrm{in}}} \din(v) \pbar_v \approx
  \sum_{v \in V_{\mathrm{in}}} \din(v) \mu_p = |E| \mu_p \label{eq:troll_mup}\quad \text{and} \\
  \sum_{u \in V_{\mathrm{out}}} \left( \sum_{v \in \NNout(u)} q_v \right) &=
  \sum_{u \in V_{\mathrm{out}}} \dout(u) \qbar_u \approx
  \sum_{u \in V_{\mathrm{out}}} \dout(u) \mu_q = |E| \mu_q \,.% \label{eq:troll_muq} \,.
\end{align*}
Therefore, letting $\tau$ be the fraction of positive edge in the graph, we have that $\tau =
\frac{\E\left(Z \right)}{|E|} \approx \lhalf \left( \mu_p + \mu_q \right)$. This means that if
$\tauhat$ is the quantity $\tau$ computed on the training, then $\tauhat$ is close to $\left(
\pbar_v + \qbar_u \right)$. Putting together all this approximations thus provide an informal
justification of the rule \eqref{eq:predictor}.

We are now ready to describe our algorithm, which follows naturally from the formula derived above.
It takes as input a training set \trainset{} drawn at random with replacement and outputs a sign
prediction for all the edges in $E\setminus \trainset$. We call it \usrule{}, which stands for Bayes
Learning Classifier based on \emph{tr}ollness and \emph{un}trustworthiness.
\iffalse
Despite its apparent simplicity, \usrule{}
works reasonably well in practice, as demonstrated by our experiments (see \autoref{s:exp}).
Moreover, unlike previous edge sign prediction methods for directed graphs, our classifier comes
with a rigorous theoretical motivation, since it approximates the Bayes optimal classifier
$y^*(u,v)$ with respect to the generative model defined in \autoref{s:gen}. It is important to point
out that when we use $1-\htr(u)$ and $1-\hun(v)$ to estimate $p_u$ and $q_v$, an additive bias shows
up due to \eqref{e:pout} and \eqref{e:pin}. This motivates the need of a threshold parameter $\tau$
to cancel this bias. Yet, the presence of a prior distribution $\mu(p,q)$ ensures that this bias is
the same for all edges $(u,v) \in E$.

@@

\textbf{4.} Any remaining non-sampled edge $(u,v)$ is predicted as $ \yhat(u,v) =
\sgn\big(\hdeltaout(u) + \hdeltain(v) - \tfrac{1}{2} - \tauhat \big) $.

Our algorithm works under the assumption that for given parameters $Q$ (a positive integer) and
$\alpha \in (0,1)$, there exists a set\footnote{$E_L$ is needed to find an estimate $\tauhat$ of
$\tau$ in~(\ref{eq:predictor}), as done in Step~3 of the algorithm. Any undirected matching of $G$ of
size $O(\log|V|)$ can be used, obtained for instance by the blossom algorithm~\autocite{matching65}.
In practice, however, we never computed $E_L$, and estimated $\tau$
on the entire training set \trainset{}.} $E_L \ss E$ of size $\tfrac{2Q}{\alpha}$ where each vertex
$u \in V$ appearing as an endpoint of some edge in $E_L$ occurs at most once as origin ---\ie{}
$(u,v)$--- and at most once as destination ---\ie{} $(v,u)$. Moreover, we assume \trainset{} has
been drawn from $E$ at random {\em without} replacement, with $m = |E_0| = \alpha\,|E|$. The
algorithm performs the following steps:
\fi

\begin{enumerate}[label=\textbf{\arabic*.}]
	\item For each $u \in V$, let $\htr(u) = \hdout^-(u)/\hdout(u)$, \ie{} the fraction of
	  negative edges found in $\Nout(u) \cap \trainset{}$.

	\item For each $v \in V$, let $\hun(v) = \hdin^-(v)/\hdin(v)$, \ie{} the fraction of
	  negative edges found in $\Nin(v) \cap \trainset{}$.

	\item Let $\tauhat$ be the fraction of positive edges in $E_L\cap \trainset{}$ (this set
	  $E_L$ is a technical requirement that we shall define shortly).

	\item Any edge $(u,v) \in E\setminus \trainset{}$ is predicted as
		$$
		\yhat(u,v) = \sgn\Big(\big(1-\htr(u)\big) + \big(1-\hun(v)\big) -\tauhat - \tfrac{1}{2}\Big)
		$$
\end{enumerate}

Despite its apparent simplicity, \usrule{}
works reasonably well in practice, as demonstrated by our experiments (see \autoref{s:exp}).
Moreover, unlike previous edge sign prediction methods for directed graphs, our classifier comes
with a rigorous theoretical motivation, since it approximates the Bayes optimal classifier
$y^*(u,v)$ with respect to the generative model defined in \autoref{s:gen}. Indeed, we quantify this
approximation on nodes whose in-degree and out-degree are not too small in the next result. However,
this first requires a few extra assumption on the training set. Namely, given parameters $Q$ (a
positive integer) and $\alpha = \nicefrac{|\trainset|}{|E|} \in (0,1)$, we assume there exists a set
$E_L \ss E$ of size $\tfrac{2Q}{\alpha}$ where each vertex $u \in V$ appearing as an endpoint of
some edge in $E_L$ occurs at most once as origin ---\ie{} \euv{}--- and at most once as destination
---\ie{} \evu{}. While the definition of $E_L$ is not immediately intuitive, this set is needed to
find an estimate $\tauhat$ of $\tau$ in~\eqref{eq:predictor} during Step~3 of \usrule{}, and its
definition allows us to applying an Hoeffding bound on independent variables. Any undirected
matching of $G$ of size $O(\log|V|)$ can be used, obtained for instance by the blossom
algorithm~\autocite{matching65}. In practice, however, we never computed $E_L$, and estimated $\tau$
on the entire training set \trainset{} (instead of $E_L \cap \trainset$).

\begin{theorem}
  \label{t:active}
  Let $G(Y) = (V,E(Y))$ be a directed graph with labels on the edges generated according to the
  model in \autoref{s:gen}.
  If the algorithm is run with parameter $Q = \Omega(\log|V|)$, and $\alpha \in (0,1)$ such that the
  above assumption about $E_L$ is satisfied, then $\yhat(u,v) = y^*(u,v)$ holds with high probability
  simultaneously for all test edges $(u,v) \in E$ such that $\dout(u),\din(v) = \Omega(\log|V|)$, and
  $\eta(u,v) = \Pr(\yuv=1)$ is bounded away from $\tfrac{1}{2}$.
\end{theorem}

While we defer the full proof of \autoref{t:active} to the additional material at the end of this
chapter (\autoref{ssec:troll_proof_bayes}), here we give a sketch of the method used.
First, by \autoref{l:bins} on negatively associated random variables (stated
\vpageref{l:bins}), we show that with our choice of $Q$, for $\theta=\frac{2Q}{\alpha}$ and for any
$u$ having enough out neighbors (\ie{} $\dout(u) \ge \theta$), at least $Q$ edges of $\Nout(u)$
are in the training set with high probability. Likewise, for any $v$ with enough in neighbors (\ie
$\din(v) \ge \theta$), at least $Q$ edges of $\Nin(v)$ are in \trainset with high probability.
With this number of samples, we then show a chain of concentration results roughly following the
outline given earlier to justify the equation~\eqref{eq:predictor}, culminating in proving that 
for any $0 < \ve < \tfrac{1}{16}$,
\[ \left| \left( 1 - \htr(u) \right) + \left( 1 - \hun(v) \right) - \tauhat
  - \frac{p_u+q_v}{2} \right| \le 8\ve \]
simultaneously holds with high probability for each $(u,v) \in \theset{(u,v) \in E}{\din(v) \ge
  \theta,\, \dout(u) \ge \theta} \setminus E_0$.

@@

The approach leading to \autoref{t:active} requires the graph to be sufficiently dense. At first sight,
having $Q = \Omega(\log|V|)$ training edges per nodes appears to be a reasonable assumption.
Consider for instance Facebook ---which is neither signed nor directed though. It has two billion
users as of 2017~\bottomfootnote{%
\url{https://investor.fb.com/investor-news/press-release-details/2017/Facebook-Reports-Second-Quarter-2017-Results/}},
each of them having $155\approx 7.2\log|V|$ friends on average~\autocite{facebookFriend16}. However,
the constant in the $\Omega$ notation is a trade off between the number of edges to sample per
node, and the quality guarantee of the $\frac{p_u+q_v}{2}$ estimation. In the Facebook example,
having a good guarantee that holds simultaneously for all the test edges might require more than
$7.2\log|V|$ samples per node. We will nonetheless see in the experiments that we can still apply
\usrule{} with satisfying results, especially since its simplicity makes it very scalable.
Additionally, and in order to sidestep this density limitation, we now introduce a second method
based on label propagation.

@@

We now proceed to solve \eqref{e:quadratic} via label propagation~\autocite{LabelPropa03} on the
graph $G''$ obtained through the second reduction of \autoref{s:prel}.
However, because of the presence of negative edge weights in $G''$, we first have to symmetrize\footnote{%
While we note here that such linear transformation of the variables does not change the problem, we
provide more details in \autoref{ssec:troll_lprop_weights} of the supplementary material.} variables
$p_i, q_i$ and $\yuv$ so as they all lie in the interval $[-1,+1]$.
After this step, one can see that, once we get back to the original variables, label propagation
computes the harmonic solution minimizing the function

### 2017-10-02 troll/generative.tex
explain better the first troll batch algo

Notice that the Bayes optimal prediction for $\yuv$ is $y^*(u,v) = \sgn\big(\eta(u,v) -
\tfrac{1}{2}\big)$, where $\eta(u,v) = \Pr\big( \yuv = 1 \big)$. Moreover, the probability of
drawing at random a $+1$-labeled edge from $\Nout(u)$ equals
\begin{equation}\label{e:pout}
	\frac{1}{2}\,\Biggl(p_u + \frac{1}{\dout(u)} \sum_{v\in \NNout(u)} q_v \Biggl) =
		\frac{1}{2}\,\left(p_u + \qbar_u\right)\,,
\end{equation}
where $\qbar_u = \frac{1}{\dout(u)} \sum_{v\in \NNout(u)} q_v$ is the average $q$ value of $u$ out
neighbors.
Similarly, the probability of drawing at random a $+1$-labeled edge from $\Nin(v)$ equals
\begin{equation}\label{e:pin}
	\frac{1}{2}\,\Biggl(q_v + \frac{1}{\din(v)} \sum_{u\in \NNin(v)} p_u \Biggl)=
		\frac{1}{2}\,\left(q_v + \pbar_v\right)
\end{equation}

>>>>

Notice that the Bayes optimal prediction for $\yuv$ is $y^*(u,v) = \sgn\big(\eta(u,v) -
\tfrac{1}{2}\big)$, where $\eta(u,v) = \Pr\big( \yuv = 1 \big)$. Moreover, once all the signs have
been generated, we can compute the expected number of positive edges outgoing from $u$ as the
expected value of the random variable $P_u = \sum_{v\in \NNout(u)} \Ind{\yuv = +1}$. Using this, we
have that the probability of
drawing at random a $+1$-labeled edge from $\Nout(u)$ equals
\begin{equation}\label{e:pout}
	\frac{\E\left(P_u\right)}{\dout(u)} =
	\frac{1}{\dout(u)} \sum_{v\in \NNout(u)} \frac{p_u + q_v}{2} =
	\frac{1}{2}\,\Biggl(p_u + \frac{1}{\dout(u)} \sum_{v\in \NNout(u)} q_v \Biggl) =
		\frac{1}{2}\,\left(p_u + \qbar_u\right)\,,
\end{equation}
where $\qbar_u = \frac{1}{\dout(u)} \left( \sum_{v\in \NNout(u)} q_v \right)$ is the average $q$
value of $u$ out neighbors.
Similarly, the probability of drawing at random a $+1$-labeled edge from $\Nin(v)$ equals
\begin{equation}\label{e:pin}
	\frac{1}{2}\,\Biggl(q_v + \frac{1}{\din(v)} \sum_{u\in \NNin(v)} p_u \Biggl)=
		\frac{1}{2}\,\left(q_v + \pbar_v\right)
\end{equation}
